Seniority and hierarchy configuration interaction for radicals and excited states

### 2023-08-20 Manuscript/manuscript.tex
hCI section
comp det

For a closed-shell reference determinant ($s_0=0$), hCI0 reduces to the \titou{restricted HF} determinant.
The first non-trivial order is hCI1 (see left panel of Fig.~\ref{fig:hCI}), which accounts for all single excitations (as CIS)
plus all paired double excitations (two electrons promoted from the same occupied orbital into the same virtual orbital),
as shown in the top left panel of Fig.~\ref{fig:determinants}.
For both classes of determinants, their number scales as $N^2$, and are thus taken into account at the same hierarchy of hCI ($h=1$ in this case).
An odd number of excitations from a closed-shell reference always leave unpaired electrons, hence the empty blocks at $s=0$ for odd $e$.
hCI1.5 augments the set of hCI1 determinants with the set of double excitations where two electrons are unpaired.
At the next integer order, hCI2 incorporates all classes of determinants where $\Ndet$ scales as $N^4$.
In total, it accounts for all single and double excitations (as CISD), plus the subset of triple excitations that leave only two unpaired electrons,
plus the subset of quadruple excitations where no electrons are unpaired.
More generally, in hCI, the number of determinants $\Ndet$ scales polynomially with the number of basis functions $N$ as $N^{2h}$.
Likewise, in eCI, $\Ndet$ scales as $N^{2e}$.
\titou{The computational scaling} along the two routes is shown in Table~\ref{tab:scaling}.

@@

For a reference containing unpaired electrons, an approximate CI model generally produces spin-contaminated states.
One can impose the solutions to have a well-defined spin by including additional determinants generated via higher-order hierarchies, which account for the missing spin-flip configurations.
The spin-contamination problem and its solution are therefore equivalent to that encountered in eCI. \cite{Maurice_1996}
Here, we have employed this procedure and considered pure spin states.
For the $s_0 = 2$ reference, one could alternatively employ a high-spin triplet determinant (two unpaired spin-up electrons), 
rather than the low-spin determinant shown in Fig.~\ref{fig:hCI}.
Regardless of the choice, hCI (as eCI) produces the same energies for the triplet states, provided that a spin eigenstate is imposed.

@@

The hCI models introduced here were implemented in {\QP} \cite{Garniron_2019} through a straightforward modification of the
\textit{configuration interaction using a perturbative selection made iteratively} (CIPSI) algorithm. \cite{Huron_1973,Giner_2013,Giner_2015,Garniron_2018}
By allowing only for the determinants that are connected with the reference determinant(s) up to a given maximum hierarchy $h$,
the CIPSI algorithm is restricted to the truncated Hilbert space specified by the reference determinant(s), and the value of $h$.
{\QP} \cite{Garniron_2019} was also employed to perform all the present eCI, sCI, and FCI calculations.
In a given calculation, the energies are considered to be converged when the (largest) EN2 correction computed in the truncated Hilbert space
lies below \SI{0.01}{\milli\hartree}. \cite{Garniron_2018}
This selected CI procedure requires considerably fewer determinants than the total number of determinants in the truncated Hilbert space
while delivering fairly converged energies.
The ground- and excited-state CI energies are obtained with the Davidson iterative algorithm. \cite{Davidson_1975}
For a given approximate CI model, we further evaluated the standard and the renormalized EN2 perturbative correction. \cite{Garniron_2019}
This calculation involves a single loop over the determinants left outside the truncated (internal) space but connected to it via at most double excitations.
Looping over these external doubly excited determinants has a computational scaling equal to $\Ndet$,
thus $\order*{N^{2e+4}}$ for eCI and $\order*{N^{2h+4}}$ for hCI, where $e$ or $h$ define the internal CI space.
For example, hCI2 and CISD present a $\order*{N^6}$ computational scaling, whereas hCI2+EN2 and CISD+EN2 scale as $\order*{N^8}$,
though with a small prefactor stemming from the EN2 calculation.
The computational scaling associated with the CI+EN2 calculations is also presented in Table~\ref{tab:scaling}.

To gauge the performance of hCI, eCI and sCI for radical systems,
we have calculated the ground-state potential energy curves (PECs) for the dissociation of four radicals:
\ce{OH}, \ce{CN}, vinyl (\ce{C2H3}), and \ce{H7}.
The CI calculations employed the ground-state restricted open-shell HF orbitals, described with the cc-pVDZ basis set, and within the frozen core approximation.
For such small systems and basis sets, FCI is attainable and provides the reference results for gauging the approximate CI models.
The equilibrium geometry of vinyl was taken from Ref.~\onlinecite{Loos_2020}, and is also reproduced in the {\SupInf}.
Their PECs were computed along the \ce{C=C} double bond breaking coordinate, with the remaining internal coordinates kept frozen.
For \ce{H7}, we considered equally spaced and linearly arranged hydrogen atoms and the PECs were computed along the symmetric dissociation coordinate.

The results were analyzed along the same lines as our previous report on hCI for closed-shell systems. \cite{Kossoski_2022}
Namely, for the different CI models considered here,
we evaluated the convergence of the non-parallelity error (NPE), the distance error, the harmonic vibrational frequencies, and the equilibrium bond lengths, as functions of $\Ndet$.
The NPE of a given level of theory is defined as the maximum minus the minimum differences between its corresponding PEC and the FCI PEC, for a given range of coordinates.
Here, we redefine the previous definition of the distance error \cite{Kossoski_2022} to accommodate for the fact the approximate PEC might appear below the FCI one when perturbative corrections are employed.
In such cases and with the previous definition, undesired negative values could be attained.
The distance error is redefined based on the signed differences between two PECs, as the absolute value of their maximum difference plus the absolute value of their minimum difference,
evaluated at a given coordinate interval.
This new definition measures how close are two PECs, remaining always non-negative.
From here on, we employ equilibrium properties when referring to both the equilibrium geometry and the harmonic vibrational frequency.
Details about how the equilibrium properties were obtained from the calculated PECs,
along with the ranges defining the NPE and distance errors can be found in the \SupInf.

The various CI models introduced here were further assessed based on calculated vertical excitation energies for 60 electronic states,
from 18 closed-shell systems and 8 radicals, with geometries extracted from the QUEST database. \cite{Veril_2021}
The full set of excited states and calculated excitation energies, for the various CI models, are provided in the {\SupInf}.
We employed the aug-cc-pVDZ basis set for systems having up to three non-hydrogen atoms and the 6-31+G(d) basis set for the larger ones.
Core orbitals were frozen.
We impose the CI solutions to be eigenstates of the spin angular momentum operator, which implies in accounting for a set of appropriate spin-flipped determinants stemming from higher-order excitation or hierarchy.
In this sense, our CIS calculations for radicals actually correspond to the so-called extended CIS \cite{Maurice_1996} for instance, and equivalently for the other hCI, eCI, and sCI models.
Notice that spin-contaminated solutions would have different energies than the spin eigenstates considered here.

We performed calculations following both the standard ground-state-based CI route and the state-specific CI route. \cite{Kossoski_2023}
For the latter, we employed the state-specific orbitals obtained in Ref.~\onlinecite{Kossoski_2023}.
Notice that, in contrast to eCI, the energies obtained with sCI and hCI models are not invariant under rotations within the occupied and virtual subspaces.
The computed excitation energies were benchmarked against the reference values of FCI quality provided in the QUEST database. \cite{Veril_2021}
The restricted HF solution (restricted open-shell HF for the open-shell systems) was taken as the reference determinant for the hCI and eCI ground state calculations
(left and middle pannels of Fig.~\ref{fig:hCI}).
For the state-specific hCI and eCI calculations, we employed a minimal CSF reference: \cite{Kossoski_2023}
a single open-shell determinant for the doublet excited states and a single open-shell CSF for the excited states from closed-shell systems
(middle and right pannels of Fig.~\ref{fig:hCI}, respectively).

As for the sCI models,
we considered sCI1 for both ground and (state-specific) excited state calculations for the open-shell systems (odd number of electrons).
For the closed-shell systems (even number of electrons), we explored two models.
In the sCI2/sCI0 model, the ground state was described with sCI0 and the excited state with sCI2, which is the minimal sCI calculation for computing singly-excited states of closed-shell systems.
Further including the $s=2$ sector for the ground state calculations gives rise to the sCI2/sCI2 model.
From here on, CI models carrying the $\Delta$ symbol denote a state-specific approach, whereas those without it correspond to a ground-state-based approach.

>>>>

For a closed-shell reference determinant ($s_0=0$), hCI0 reduces to the \titou{restricted HF} determinant.
The first non-trivial order is hCI1 (see left panel of Fig.~\ref{fig:hCI}), which accounts for all single excitations (as CIS)
plus all paired double excitations (two electrons promoted from the same occupied orbital into the same virtual orbital),
as shown in the top left panel of Fig.~\ref{fig:determinants}.
For both classes of determinants, their number scales as $N^2$, and are thus taken into account at the same hierarchy of hCI ($h=1$ in this case).
An odd number of excitations from a closed-shell reference always leave unpaired electrons, hence the empty blocks at $s=0$ for odd $e$.
hCI1.5 augments the set of hCI1 determinants with the set of double excitations where two electrons are unpaired.
At the next integer order, hCI2 incorporates all classes of determinants where $\Ndet$ scales as $N^4$.
In total, it accounts for all single and double excitations (as CISD), plus the subset of triple excitations that leave only two unpaired electrons,
plus the subset of quadruple excitations where no electrons are unpaired.
More generally, in hCI, the number of determinants $\Ndet$ scales polynomially with the number of basis functions $N$ as $N^{2h}$.
Likewise, in eCI, $\Ndet$ scales as $N^{2e}$.
\titou{The computational scaling} along the two routes is shown in Table~\ref{tab:scaling}.
\PFL{Should we mention something about the scaling of seniority-based CI here?}

@@

For a reference containing unpaired electrons, an approximate CI model generally produces spin-contaminated states.
One can impose the solutions to have a well-defined spin by including additional determinants generated via higher-order hierarchies, which account for the missing spin-flip configurations. \cite{Chilkuri_2021}
The spin-contamination problem and its solution are therefore equivalent to that encountered in eCI. \cite{Maurice_1996}
Here, we have employed this procedure and considered pure spin states.
For the $s_0 = 2$ reference, one could alternatively employ a high-spin triplet determinant (two unpaired spin-up electrons), 
rather than the low-spin determinant shown in Fig.~\ref{fig:hCI}.
Regardless of the choice, hCI (as eCI) produces the same energies for the triplet states, provided that a spin eigenstate is imposed.

@@

The hCI models introduced here were implemented in {\QP} \cite{Garniron_2019} through a straightforward modification of the
\textit{configuration interaction using a perturbative selection made iteratively} (CIPSI) algorithm. \cite{Huron_1973,Giner_2013,Giner_2015,Garniron_2018}
By allowing only for the determinants that are connected with the reference determinant(s) up to a given maximum hierarchy $h$,
the CIPSI algorithm is restricted to the truncated Hilbert space specified by the reference determinant(s), and the value of $h$.
{\QP} \cite{Garniron_2019} was also employed to perform all the present eCI, sCI, and FCI calculations.
In a given calculation, the energies are considered to be converged when the (largest) EN2 correction computed in the truncated Hilbert space
lies below \SI{0.01}{\milli\hartree}. \cite{Garniron_2018}
This selected CI procedure requires considerably fewer determinants than the total number of determinants in the truncated Hilbert space
while delivering fairly converged energies.
The ground- and excited-state CI energies are obtained with the Davidson iterative algorithm. \cite{Davidson_1975}
For a given approximate CI model, we further evaluated the standard and the renormalized EN2 perturbative correction. \cite{Garniron_2019}
This calculation involves a single loop over the determinants left outside the truncated (internal) space but connected to it via at most double excitations.
Looping over these external doubly excited determinants has a computational scaling equal to $\Ndet$,
thus $\order*{N^{2e+4}}$ for eCI and $\order*{N^{2h+4}}$ for hCI, where $e$ or $h$ define the internal CI space.
For example, hCI2 and CISD present a $\order*{N^6}$ computational scaling, whereas hCI2+EN2 and CISD+EN2 scale as $\order*{N^8}$,
though with a small prefactor stemming from the EN2 calculation that employs a very efficient semistochastic algorithm. \cite{Garniron_2018}
The computational scaling associated with the CI+EN2 calculations is also presented in Table~\ref{tab:scaling}.

To gauge the performance of hCI, eCI, and sCI for radicals, we have calculated the ground-state potential energy curves (PECs) for the dissociation of four radicals:
\ce{OH}, \ce{CN}, vinyl (\ce{C2H3}), and \ce{H7}.
The CI calculations employed the ground-state restricted open-shell HF orbitals, described with the cc-pVDZ basis set, and within the frozen-core approximation.
For such small systems and basis sets, FCI is attainable and provides the reference results for gauging the approximate CI models.
The equilibrium geometry of vinyl was taken from Ref.~\onlinecite{Loos_2020} and is also reproduced in the {\SupInf}.
Their PECs were computed along the \ce{C=C} double bond breaking coordinate, with the remaining internal coordinates kept frozen.
For \ce{H7}, we considered equally spaced and linearly arranged hydrogen atoms and the PECs were computed along the symmetric dissociation coordinate.

The results were analyzed along the same lines as our previous report on hCI for closed-shell systems. \cite{Kossoski_2022}
Namely, for the different CI models considered here,
we evaluated the convergence of the non-parallelity error (NPE), the distance error, the harmonic vibrational frequencies, and the equilibrium bond lengths, as functions of $\Ndet$.
The NPE of a given level of theory is defined as the maximum minus the minimum energy differences between its corresponding PEC and the FCI PEC, for a given range of coordinates.
Here, we redefine the previous definition of the distance error \cite{Kossoski_2022} to accommodate for the fact the approximate PEC might appear below the FCI one when perturbative corrections are employed.
In such cases and with the previous definition, undesired negative values could be attained.
The distance error is redefined based on the signed differences between two PECs, as the absolute value of their maximum difference plus the absolute value of their minimum difference,
evaluated at a given coordinate interval.
This new definition measures how close are two PECs, remaining always non-negative.
From here on, we employ equilibrium properties when referring to both the equilibrium geometry and the harmonic vibrational frequency.
Details about how the equilibrium properties were obtained from the calculated PECs,
along with the ranges defining the NPE and distance errors can be found in the \SupInf.

The various CI models introduced here were further assessed based on calculated vertical excitation energies for 60 electronic states,
from 18 closed-shell systems and 8 radicals, with geometries extracted from the QUEST database. \cite{Veril_2021}
The full set of excited states and calculated excitation energies, for the various CI models, are provided in the {\SupInf}.
We employed the aug-cc-pVDZ basis set for systems having up to three non-hydrogen atoms and the 6-31+G(d) basis set for the larger ones.
Core orbitals were frozen systematically.
We impose the CI solutions to be eigenstates of the spin angular momentum operator, which implies accounting for a set of appropriate spin-flipped determinants stemming from higher-order excitations or hierarchies. \cite{Chilkuri_2021}
In this sense, our CIS calculations for radicals actually correspond to the so-called extended CIS \cite{Maurice_1996} for instance, and equivalently for the other hCI, eCI, and sCI models.
Notice that spin-contaminated solutions would have different energies than the spin eigenstates considered here.

We performed calculations following both the standard ground-state-based CI route and the state-specific CI route. \cite{Kossoski_2023}
For the latter, we employed the state-specific orbitals obtained in Ref.~\onlinecite{Kossoski_2023}.
Notice that, in contrast to eCI, the energies obtained with sCI and hCI models are not invariant under rotations within the occupied and virtual subspaces.
The computed excitation energies were benchmarked against the reference values of FCI quality provided in the QUEST database. \cite{Veril_2021}
The restricted HF solution (restricted open-shell HF for the open-shell systems) was taken as the reference determinant for the hCI and eCI ground-state calculations
(left and middle panels of Fig.~\ref{fig:hCI}).
For the state-specific hCI and eCI calculations, we employed a minimal CSF reference: \cite{Kossoski_2023}
a single open-shell determinant for the doublet excited states and a single open-shell CSF for the excited states from closed-shell systems
(middle and right panels of Fig.~\ref{fig:hCI}, respectively).

As for the sCI models,
we considered sCI1 for both ground and (state-specific) excited-state calculations for the open-shell systems (odd numbers of electrons).
For the closed-shell systems (\titou{even numbers of electrons}), we explored two models.
In the sCI2/sCI0 model, the ground state was described with sCI0 and the excited state with sCI2, which is the minimal sCI calculation for computing singly-excited states of closed-shell systems.
Further including the seniority-two sector for the ground-state calculations defines the sCI2/sCI2 model.
From here on, CI models carrying the $\Delta$ symbol denote a state-specific approach, whereas those without it correspond to a ground-state-based approach.
